---
title: "ST 558 Project 1"
format: html
author: Cameron Mullaney
date: today
editor: visual
---

# Setup

Loading in some packages I'll need, as well as my API key

```{r}
#| warning: false
library(httr)
library(jsonlite)
library(tidyverse)
library(lubridate)

APIkey <- "3e4e9dd9b93b8546b5de301ae8c11506545540b9"

```

# Part I: Data Processing

## Parsing Variable Values

### Variables Query Function

This function uses the PUMS API to pull in the list of census variables from a given year, as well as features of those variables (range, data type, etc.)

```{r}
pums <- function(url){
  ## First, check year of URL to determine column names
  if (grepl("2021|2022", url)){
    print("2021-2")
    colList <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER",
               "HHL", "SCH", "SCHL", "SEX", "ST", "REGION", "DIVISION")
  } else {
    print("2023-4")
    colList <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP", "FER",
                 "HHL", "SCH", "SCHL", "SEX", "STATE", "REGION", "DIVISION")
  }
  
  ## Then, pull in variable info, parse to JSON, and return subsetted list
  getJson <- GET(url)
  parsed <- fromJSON(rawToChar(getJson$content))
  vars <- as.list(parsed$variables)
  
  varSubset <- vars[colList]
  
  return(varSubset)
}
```

Running `pums()` and saving results to .rds file in repository

```{r}
#| results: hold
data2021 <- pums("https://api.census.gov/data/2021/acs/acs1/pums/variables.json")
data2022 <- pums("https://api.census.gov/data/2022/acs/acs1/pums/variables.json")
data2023 <- pums("https://api.census.gov/data/2023/acs/acs1/pums/variables.json")
data2024 <- pums("https://api.census.gov/data/2024/acs/acs1/pums/variables.json")
```

```{r}
saveRDS(list(pums2021 = data2021, pums2022 = data2022, 
             pums2023 = data2023, pums2024 = data2024), file = "data/pumsVars.rds")
```

## Parsing API Data

### Functions

#### Function 1: Query to tibble (helper)

```{r}
tibblehelper <- function(rawinput){
  a <- fromJSON(rawToChar(rawinput$content))
  as.data.frame(a[-1, , drop = FALSE], stringsAsFactors = FALSE) |>
    setNames(a[1, ]) |>
    as_tibble()
}
```

#### Function 2: Time parsing (helper)

*For this sub-section an LLM was used to assist in function writing.*

```{r}
time_calc <- function(label) {
  '
  This function will return either the midpoint of a provided time range.
  If the provided label is missing a time, an NA time is returned
  '
  ### Runs once per element
  label <- as.character(label)
  u <- unique(label)
  
  mids <- map_dbl(u, function(x) {
    
    if (is.na(x) || !str_detect(x, " to ")) return(NA_real_)
    
    ## Split time range into two values, make am/pm inidcators uniform
    
    parts <- str_split_1(x, " to ") |>
      str_replace_all("a\\.m\\.", "AM") |>
      str_replace_all("p\\.m\\.", "PM")
    
    ## Parse each piece into date/time, check for NA here too
    
    t <- parse_date_time(parts, orders = c("I:M p", "I p"), quiet = TRUE)
    if (any(is.na(t))) return(NA_real_)
    
    ## Check if the interval goes past midnight, adjust accordingly
    
    if (t[2] < t[1]) t[2] <- t[2] + days(1)  
    
    ## Compute midpoint, and convert to proper format
    
    mid <- t[1] + (t[2] - t[1]) / 2
    as.numeric(difftime(mid, floor_date(t[1], "day"), units = "secs")) %% 86400
  })
  
  hms::as_hms(mids[match(label, u)])
}
```

#### Big parsing function

This function will query the PUMS API, and allows a user to specify what variables they want to pull in. In addition, the user must supply an API key. Both `tibblehelper()` and `time_calc()` are used. The .rds file from earlier is used to pull in the proper variable values, and properly label the final tibble output.

```{r}
pums2 <- function(year = 2024,
                  num_vars = c("AGEP", "PWGTP"),
                  cat_vars = "SEX",
                  geo_level = "state",
                  geo_value = "01",
                  API_key = "") {
  
  
  levels_list <- readRDS("data/pumsVars.rds")
  
  ## Lists of acceptable column names
  
  numlist <- c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP")
  catlist <- c("FER", "HHL", "SCH", "SCHL", "SEX")
  num_vars <- union("PWGTP", num_vars)
  
  ### Check column names
  
  if (!all(num_vars %in% numlist)) stop("Invalid numeric variable")
  if (!all(cat_vars %in% catlist)) stop("Invalid categorical variable")
  
  if (length(setdiff(num_vars, "PWGTP")) < 1)
    stop("Need at least one numeric variable besides PWGTP")
  
  if (length(cat_vars) < 1) stop("Need at least one categorical variable")
  
  ## Check for proper year
  
  if (length(year) != 1 || !year %in% 2021:2024)
    stop("Must provide a single year within 2021-2024")
  
  ## Check for API key
  if (API_key == "")
    stop("No API key provided")
  
  ## Check geography level
  if (length(geo_level) != 1 || !geo_level %in% c("state", "division", "region"))
    stop("geo_level must be state, division, or region")
  
  ## Check geo value, set geo_api based on geo_value input
  if (length(geo_value) != 1) stop("geo_value must be a single value (or 'all')")
  geo_api <- if (tolower(geo_value) == "all") "*" else geo_value
  
  ## Finally! Query time
  
  querypull <- GET(paste0("https://api.census.gov/data/", year, "/acs/acs1/pums"),
              query = list(get = paste(c(num_vars, cat_vars), collapse = ","),
                           `for` = paste0(geo_level, ":", geo_api),
                           key = API_key))
  out <- tibblehelper(querypull)
  
  ## Separate numeric cols into plain numeric cols and time cols
  ## Applies as.numeric to the correct columns
  
  plain_numbers <- intersect(num_vars, c("AGEP", "GASP", "GRPIP", "JWMNP", "PWGTP"))
  out <- out |> mutate(across(all_of(plain_numbers), as.numeric))
  
  ## Time variables: 
  
  for (v in intersect(num_vars, c("JWAP", "JWDP"))) {
    
    lk <- unlist(levels_list[[paste0("pums", year)]][[v]][["values"]][["item"]])
    lab <- lk[match(as.integer(out[[v]]), as.integer(names(lk)))]
    out[[v]] <- time_calc(unname(lab))
  }
  
  ## Categorical variables -> factors with labels from the .rds
  for (v in cat_vars) {
    
    lv <- levels_list[[paste0("pums", as.character(year))]][[v]][["values"]][["item"]]
    out[[v]] <- factor(out[[v]], levels = names(lv), labels = unname(lv))
  }

  class(out) <- c("census", class(out))
  out  
  }
```

Trying it out! More columns slows it down, but it works!

```{r}
attempt <- pums2(year = 2022, 
                 num_vars = c("AGEP", "GASP", "GRPIP", "JWAP", "JWDP", "JWMNP", "PWGTP"), 
                 cat_vars = c("FER", "HHL", "SCH", "SCHL", "SEX"), 
                 geo_level = "region", 
                 geo_value = "all", 
                 API_key = APIkey)
```

#### Combining years

This function will loop `pums2()` over a list of years provided, producing a final tibble with data from multiple years.

```{r}

combo <- function(years = c(2021:2024),
                  num_vars = c("AGEP", "PWGTP"),
                  cat_vars = "SEX",
                  geo_level = "state",
                  geo_value = "01",
                  API_key = ""){
  
  ## Initialize empty tibble
  bigList <- tibble()
  
  cat("year \t rows\n")
  for (i in years){
    temp <- pums2(year = i, 
                  num_vars = num_vars, 
                  cat_vars = cat_vars,
                  geo_level = geo_level,
                  geo_value = geo_value,
                  API_key = API_key)
    
    ## Add year column
    temp <- temp |> add_column(year = i)
    
    bigList <- bind_rows(bigList, temp)
    
    ## Print number of rows in each year
    cat(i, "\t", nrow(temp), "\n")
  }
  
  return(bigList)
  
}
```

Testing it out - Functional!

```{r}
test <- combo(years = c(2021: 2023), API_key = APIkey)

```

# Part II: Writing a Generic Function for Summarizing
